---
title: 'NMF-RE Simulation: Baseline (P=4, Q=1)'
author: 'Kenichi Satoh'
date: "Last updated: `r format(Sys.time(), '%Y/%m/%d')`"
output:
  word_document:
    fig_width: 12
    fig_height: 12
    toc: no
---

This script reproduces Table 1 (and Appendix Table F1) of
"Wild Bootstrap Inference for Non-Negative Matrix Factorization
with Random Effects" (Satoh, 2026).

Baseline simulation: P=4, Q=1, Orthodont-based DGP.
Two error distributions (Gaussian, Exp-centered) x
two hypotheses (Null, Alternative) x three sample sizes (N=27, 100, 200).
R=1000 Monte Carlo replicates per setting.

NOTE: This simulation is computationally intensive.
With 24 cores, expect ~2-3 hours for the full run.

## Load NMF-RE core functions
```{r, dpi=300, eval=TRUE, echo=TRUE, warning=TRUE, error=TRUE, message=TRUE}
source("nmfre_core.R")

############################################################
## NMF-RE Numerical Study (Orthodont, P=4, Q=1)  [ALIGNED+]
## - Uses the SAME nmf.re() as the data-analysis example
## - Adds option: err_dist ("gaussian" or "exp_centered")
############################################################

Sys.setenv(
  OMP_NUM_THREADS="1",
  OPENBLAS_NUM_THREADS="1",
  MKL_NUM_THREADS="1",
  VECLIB_MAXIMUM_THREADS="1",
  NUMEXPR_NUM_THREADS="1"
)

suppressPackageStartupMessages({
  library(nlme)
  library(parallel)
  # install.packages("nmfkc")  # required for nmfkc::nmfkc() initialization
})

## ==========================
## Settings
## ==========================
N_vec   <- c(27, 100, 200)
R_mc    <- 1000
wild_B  <- 1000

## ==========================
## Full (refit) bootstrap settings
## Set full_B = 0 to skip (very time-consuming)
## ==========================
full_B      <- 0
full_level  <- 0.95
full_weight <- "exp_centered"
full_maxit  <- 50000
full_eps    <- 1e-8
seed_base_full <- 20260301

n_cores <- max(1, detectCores() - 1)

seed_base_design <- 20260110
seed_base_mc     <- 20260201

Q <- 1
maxit   <- 50000
epsilon <- 1e-8

dfU_cap_rate <- 0.21
alpha <- 0.05

sigma2_true <- 1
tau2_true   <- 1

err_dist_vec <- c("gaussian", "exp_centered")
u_dist       <- "gaussian"

## True Orthodont-based structure
X_true <- matrix(c(0.2308, 0.2409, 0.2566, 0.2717), nrow=4, ncol=1)
stopifnot(abs(sum(X_true) - 1) < 1e-6)

Theta_true_B <- matrix(c(90.502, 9.4285), nrow=1,
                       dimnames=list("Trend1", c("intercept","male")))
Theta_true_A <- Theta_true_B
Theta_true_A[1,"male"] <- 0

## ==========================
## Design matrix A
## ==========================
dat8 <- subset(nlme::Orthodont, age == 8)
male0 <- as.integer(dat8$Sex == "Male")
p_male <- mean(male0)

A_orth <- rbind(intercept = rep(1, length(male0)),
                male      = male0)
stopifnot(ncol(A_orth) == 27)

make_A_bern <- function(N, p, seed) {
  set.seed(seed)
  male <- rbinom(N, 1, p)
  rbind(intercept = rep(1, N),
        male      = male)
}

use_orthodont_design_N27 <- TRUE

get_A <- function(N_target) {
  if (N_target == 27 && isTRUE(use_orthodont_design_N27)) {
    return(A_orth)
  }
  make_A_bern(N_target, p_male, seed_base_design + N_target)
}

## ==========================
## Noise generators
## ==========================
gen_noise_vec <- function(n, dist, sd) {
  dist <- match.arg(dist, c("gaussian", "exp_centered"))
  if (!is.finite(sd) || sd < 0) stop("sd must be nonnegative.")
  if (sd == 0) return(rep(0, n))

  if (dist == "gaussian") {
    return(rnorm(n, mean = 0, sd = sd))
  } else {
    return((rexp(n, rate = 1) - 1) * sd)
  }
}

gen_noise_mat <- function(nrow, ncol, dist, sd) {
  matrix(gen_noise_vec(nrow * ncol, dist = dist, sd = sd), nrow = nrow, ncol = ncol)
}

## ==========================
## One replicate
## ==========================
simulate_one <- function(sim_id, N_target, Theta_true, scenario_label,
                         err_dist="gaussian", u_dist="gaussian") {

  A_use <- get_A(N_target)

  seed_mc <- seed_base_mc + sim_id + 100000 * N_target
  set.seed(seed_mc)

  U <- matrix(gen_noise_vec(ncol(A_use), dist=u_dist, sd=sqrt(tau2_true)), 1)
  E <- gen_noise_mat(nrow(X_true), ncol(A_use), dist=err_dist, sd=sqrt(sigma2_true))
  B <- Theta_true %*% A_use + U
  Y <- X_true %*% B + E

  ymin <- suppressWarnings(min(Y))
  if (is.finite(ymin) && ymin < 0) Y <- Y - ymin + 1e-8

  ini <- tryCatch(
    nmfkc::nmfkc(
      Y = Y, A = A_use, Q = Q,
      method = "EU",
      epsilon = 1e-8,
      maxit = 2000,
      seed = 700000 + sim_id,
      print.trace = FALSE,
      print.dims  = FALSE,
      save.time = FALSE,
      save.memory = FALSE
    ),
    error = function(e) e
  )
  if (inherits(ini, "error")) return(NULL)
  if (is.null(ini$X) || is.null(ini$C)) return(NULL)

  X_ini <- as.matrix(ini$X)
  Theta_ini <- as.matrix(ini$C)

  s <- colSums(X_ini)
  s[!is.finite(s) | s <= 0] <- 1
  X_ini <- sweep(X_ini, 2, s, "/")
  Theta_ini <- sweep(Theta_ini, 1, s, "*")
  Theta_ini <- pmax(Theta_ini, 1e-6)

  fit <- tryCatch(
    nmf.re(
      Y = Y, A = A_use, Q = Q,
      X.ini = X_ini,
      Theta.ini = Theta_ini,
      prefix = "Trend",
      verbose = FALSE,
      dfU_control = "cap",
      dfU_cap_rate = dfU_cap_rate,
      maxit = maxit,
      epsilon = epsilon,
      inference = TRUE,
      wild_B = wild_B,
      wild_weight = "exp_centered",
      wild_seed = 100000 + sim_id,
      seed = 500000 + sim_id,
      post_message = FALSE,
      save_obj_trace = FALSE,
      theta_p_side = "one.sided"
    ),
    error = function(e) e
  )
  if (inherits(fit, "error")) return(NULL)

  coef <- fit$coefficients
  if (is.null(coef) || nrow(coef) == 0) return(NULL)

  pick <- function(varname) {
    ii <- which(coef$Variable == varname & coef$Basis == "Trend1")
    if (length(ii) != 1) return(list(est=NA,se=NA,bse=NA))
    list(est = coef$Estimate[ii],
         se  = coef$SE[ii],
         bse = coef$BSE[ii])
  }

  mal <- pick("male")

  z_se  <- mal$est / mal$se
  z_bse <- mal$est / mal$bse

  reject_se  <- if (scenario_label == "A_null") (z_se  > qnorm(1 - alpha)) else (abs(z_se)  > qnorm(1 - alpha/2))
  reject_bse <- if (scenario_label == "A_null") (z_bse > qnorm(1 - alpha)) else (abs(z_bse) > qnorm(1 - alpha/2))

  ciL_se  <- mal$est - qnorm(1 - alpha/2) * mal$se
  ciU_se  <- mal$est + qnorm(1 - alpha/2) * mal$se
  ciL_bse <- mal$est - qnorm(1 - alpha/2) * mal$bse
  ciU_bse <- mal$est + qnorm(1 - alpha/2) * mal$bse

  theta_true_male <- Theta_true[1,"male"]

  cover_se  <- (theta_true_male >= ciL_se  && theta_true_male <= ciU_se)
  cover_bse <- (theta_true_male >= ciL_bse && theta_true_male <= ciU_bse)

  pct_cover <- NA
  if (!is.null(fit$Theta.ci.lower) && !is.null(fit$Theta.ci.upper)) {
    lo <- fit$Theta.ci.lower[1,"male"]
    hi <- fit$Theta.ci.upper[1,"male"]
    pct_cover <- (theta_true_male >= lo && theta_true_male <= hi)
  }

  ## Full (refit) bootstrap
  fbse_male <- NA_real_
  reject_fbse <- NA_integer_
  cover_fbse  <- NA_integer_
  pct_cover_full <- NA_integer_
  full_ok <- NA_integer_

  if (is.finite(full_B) && full_B > 0) {
    fb <- full_refit_bootstrap(
      Y = Y, A_use = A_use, fit0 = fit,
      sim_id = sim_id, N_target = N_target,
      B = full_B, level = full_level, weight = full_weight,
      Q = Q, dfU_cap_rate = dfU_cap_rate,
      maxit_boot = full_maxit, eps_boot = full_eps,
      seed_base_full = seed_base_full
    )
    full_ok <- as.integer(fb$ok)
    fbse_male <- fb$sd

    if (is.finite(fbse_male) && fbse_male > 0) {
      z_fbse <- mal$est / fbse_male
      reject_fbse <- if (scenario_label == "A_null") {
        as.integer(z_fbse > qnorm(1 - alpha))
      } else {
        as.integer(abs(z_fbse) > qnorm(1 - alpha/2))
      }
      ciL_fbse <- mal$est - qnorm(1 - alpha/2) * fbse_male
      ciU_fbse <- mal$est + qnorm(1 - alpha/2) * fbse_male
      cover_fbse <- as.integer(theta_true_male >= ciL_fbse && theta_true_male <= ciU_fbse)
    }
    if (is.finite(fb$lo) && is.finite(fb$hi)) {
      pct_cover_full <- as.integer(theta_true_male >= fb$lo && theta_true_male <= fb$hi)
    }
  }

  seed_mc <- seed_base_mc + sim_id + 100000 * N_target

  data.frame(
    sim = sim_id, seed_mc = seed_mc,
    y_sum = sum(Y), y_min = min(Y),
    N = N_target, scenario = scenario_label,
    err_dist = err_dist, u_dist = u_dist,
    est_male = mal$est, se_male = mal$se, bse_male = mal$bse,
    reject_se  = as.integer(reject_se),
    reject_bse = as.integer(reject_bse),
    cover_se   = as.integer(cover_se),
    cover_bse  = as.integer(cover_bse),
    pct_cover  = as.integer(pct_cover),
    dfU_frac   = fit$dfU_frac,
    fbse_male = fbse_male,
    reject_fbse = reject_fbse,
    cover_fbse  = cover_fbse,
    pct_cover_full = pct_cover_full,
    full_ok = full_ok,
    stringsAsFactors = FALSE
  )
}

run_one_setting <- function(N_target, Theta_true, scenario_label, err_dist, u_dist) {
  out <- mclapply(
    X = seq_len(R_mc),
    FUN = function(i) simulate_one(i, N_target, Theta_true, scenario_label,
                                   err_dist = err_dist, u_dist = u_dist),
    mc.cores = n_cores
  )
  ok_flag <- !vapply(out, is.null, logical(1))
  n_total <- length(out)
  n_ok <- sum(ok_flag)
  cat(sprintf("  [OK] %d / %d\n", n_ok, n_total))
  if (n_ok < 0.9 * n_total) {
    stop(sprintf("Too many failures: only %d/%d succeeded.", n_ok, n_total))
  }
  out <- out[ok_flag]
  do.call(rbind, out)
}

summarize_setting <- function(df, Theta_true) {
  theta_true_male <- Theta_true[1,"male"]
  est <- df$est_male
  sd_mc <- sd(est)
  bias  <- mean(est) - theta_true_male
  rmse  <- sqrt(mean((est - theta_true_male)^2))

  c(
    MeanFBSE = mean(df$fbse_male, na.rm = TRUE),
    RejectFBSE = mean(df$reject_fbse, na.rm = TRUE),
    CoverFBSE  = mean(df$cover_fbse,  na.rm = TRUE),
    PctCoverFull = mean(df$pct_cover_full, na.rm = TRUE),
    FullOK = mean(df$full_ok, na.rm = TRUE),
    Bias = bias, SD = sd_mc, RMSE = rmse,
    MeanSE  = mean(df$se_male,  na.rm=TRUE),
    MeanBSE = mean(df$bse_male, na.rm=TRUE),
    RejectSE  = mean(df$reject_se,  na.rm=TRUE),
    RejectBSE = mean(df$reject_bse, na.rm=TRUE),
    CoverSE   = mean(df$cover_se,   na.rm=TRUE),
    CoverBSE  = mean(df$cover_bse,  na.rm=TRUE),
    PctCover  = mean(df$pct_cover,  na.rm=TRUE),
    dfU       = mean(df$dfU_frac,   na.rm=TRUE)
  )
}

```


## Full refit bootstrap function
```{r, dpi=300, eval=TRUE, echo=TRUE, warning=TRUE, error=TRUE, message=TRUE}
make_Y_wild_resid <- function(Y, fitted, weight = c("exp_centered"), shift_nonneg = TRUE) {
  weight <- match.arg(weight)
  P <- nrow(Y); N <- ncol(Y)
  R <- Y - fitted

  if (weight == "exp_centered") {
    w <- rexp(N, rate = 1) - 1
  } else {
    stop("Unsupported weight.")
  }

  Yb <- fitted + sweep(R, 2, w, `*`)

  if (isTRUE(shift_nonneg)) {
    ymin <- suppressWarnings(min(Yb))
    if (is.finite(ymin) && ymin < 0) Yb <- Yb - ymin + 1e-8
  }
  Yb
}

full_refit_bootstrap <- function(
  Y, A_use, fit0,
  sim_id, N_target,
  B = 200, level = 0.95, weight = "exp_centered",
  Q = 1, dfU_cap_rate = 0.21,
  maxit_boot = 50000, eps_boot = 1e-8,
  seed_base_full = 20260301
) {
  if (!is.finite(B) || B <= 0) {
    return(list(theta_boot = numeric(0), ok = 0L, sd = NA_real_, lo = NA_real_, hi = NA_real_))
  }

  fitted0 <- fit0$XB.blup
  if (is.null(fitted0)) {
    fitted0 <- fit0$X %*% ((fit0$Theta %*% A_use) + fit0$U)
  }

  theta_boot <- rep(NA_real_, B)

  for (b in 1:B) {
    set.seed(seed_base_full + sim_id + 100000 * N_target + 1000 * b)

    Yb <- make_Y_wild_resid(Y, fitted = fitted0, weight = weight, shift_nonneg = TRUE)

    fb <- tryCatch(
      nmf.re(
        Y = Yb, A = A_use, Q = Q,
        X.ini = fit0$X, Theta.ini = fit0$Theta, U.ini = fit0$U,
        verbose = FALSE,
        dfU_control = "cap", dfU_cap_rate = dfU_cap_rate,
        maxit = maxit_boot, epsilon = eps_boot,
        inference = FALSE,
        seed = 500000 + sim_id + 1000 * b,
        post_message = FALSE, save_obj_trace = FALSE
      ),
      error = function(e) NULL
    )
    if (is.null(fb) || is.null(fb$Theta)) next
    th <- tryCatch(as.numeric(fb$Theta[1, "male"]), error = function(e) NA_real_)
    theta_boot[b] <- th
  }

  ok <- sum(is.finite(theta_boot))
  if (ok < max(10, 0.7 * B)) {
    return(list(theta_boot = theta_boot, ok = ok, sd = NA_real_, lo = NA_real_, hi = NA_real_))
  }

  alpha <- 1 - level
  sd_b <- sd(theta_boot, na.rm = TRUE)
  lo <- as.numeric(quantile(theta_boot, probs = alpha / 2, na.rm = TRUE, names = FALSE))
  hi <- as.numeric(quantile(theta_boot, probs = 1 - alpha / 2, na.rm = TRUE, names = FALSE))

  list(theta_boot = theta_boot, ok = ok, sd = sd_b, lo = lo, hi = hi)
}
```


## Run all settings
```{r, dpi=300, eval=TRUE, echo=TRUE, warning=TRUE, error=TRUE, message=TRUE}
all_rows <- list()

for (err_dist in err_dist_vec) {
  cat("\n==============================\n")
  cat("Running err_dist =", err_dist, " (u_dist =", u_dist, ")\n")
  cat("==============================\n")

  for (N_target in N_vec) {

    ## Null hypothesis (Theta_male = 0)
    cat("\n--- N =", N_target, "Null ---\n")
    dfA <- run_one_setting(N_target, Theta_true_A, "A_null", err_dist, u_dist)
    sumA <- summarize_setting(dfA, Theta_true_A)

    ## Alternative hypothesis (Theta_male = 9.4285)
    cat("\n--- N =", N_target, "Alternative ---\n")
    dfB <- run_one_setting(N_target, Theta_true_B, "B_alt", err_dist, u_dist)
    sumB <- summarize_setting(dfB, Theta_true_B)

    all_rows[[length(all_rows)+1]] <- data.frame(
      err_dist = err_dist, u_dist = u_dist,
      N = N_target, Scenario = "A_null", t(sumA))
    all_rows[[length(all_rows)+1]] <- data.frame(
      err_dist = err_dist, u_dist = u_dist,
      N = N_target, Scenario = "B_alt", t(sumB))
  }
}

tab <- do.call(rbind, all_rows)
print(tab)

write.csv(tab, file = "sim_summary_baseline.csv", row.names = FALSE)
cat("Saved: sim_summary_baseline.csv\n")

```

